---
title: "Data Analysis with istatistikPaketim"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Data Analysis with istatistikPaketim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, include=FALSE}
library(istatistikPaketim)
library(knitr)
library(kableExtra)
library(ggplot2)

load("../data/ogrenci_notlari.rda")
load("../data/temiz_penguinler.rda")
```


## Introduction

This vignette demonstrates how to use the `istatistikPaketim` package for basic data analysis.

The package includes datasets and functions for statistical analysis, visualization, and outlier detection.

This guide aims to show how the package can be used in a simple and practical way.

> A good vignette does not only show code, but also helps users understand the idea.

## Installation

To use the `istatistikPaketim` package, it must first be installed from GitHub.

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("nkhnurkholifah-maker/istatistikPaketim")
library(istatistikPaketim)
```

## Package Contents

The `istatistikPaketim` package contains two example datasets and two main functions.

The first dataset is `ogrenci_notlari`. This dataset contains students' mathematics, physics, and chemistry scores. It is used for basic statistical analysis such as mean, median, t-test, and outlier detection.

The second dataset is `temiz_penguinler`. This dataset contains cleaned penguin data and can be used for simple data exploration and visualization.

The package also includes two outlier detection functions. The function `aykiri_deger_analizi()` is used to detect outliers and can also show a warning when the input data is not suitable for analysis. The function `tespit_et_outlier()` is another simple function for finding outlier values in numeric data.

Before using these functions, users should make sure that the input is numeric. If text or non-numeric data is entered, the function may give a warning or error.


## Basic Statistical Concepts

In statistics, several basic measures are commonly used to understand data. These include mean, median, and standard deviation.

The **mean (average)** shows the central value of the data.

The mean can be written as $\bar{x} = \frac{\sum x_i}{n}$.

$$
\bar{x} = \frac{x_1 + x_2 + \cdots + x_n}{n}
$$

The **median** is the middle value when the data is sorted.

The **standard deviation** shows how spread out the values are from the mean.

Another important concept is the **Interquartile Range (IQR)**, which is used to detect outliers.

$$
IQR = Q3 - Q1
$$


## Explore the Dataset


```{r}
rata_matematik <- mean(ogrenci_notlari$matematik_notu)
rata_fizik <- mean(ogrenci_notlari$fizik_notu)
rata_kimya <- mean(ogrenci_notlari$kimya_notu)

ogrenci_notlari_style <- ogrenci_notlari

ogrenci_notlari_style$matematik_notu <- kableExtra::cell_spec(
  ogrenci_notlari$matematik_notu,
  format = "html",
  color = ifelse(ogrenci_notlari$matematik_notu < rata_matematik, "red", "green"),
  bold = TRUE
)

ogrenci_notlari_style$fizik_notu <- kableExtra::cell_spec(
  ogrenci_notlari$fizik_notu,
  format = "html",
  color = ifelse(ogrenci_notlari$fizik_notu < rata_fizik, "red", "green"),
  bold = TRUE
)

ogrenci_notlari_style$kimya_notu <- kableExtra::cell_spec(
  ogrenci_notlari$kimya_notu,
  format = "html",
  color = ifelse(ogrenci_notlari$kimya_notu < rata_kimya, "red", "green"),
  bold = TRUE
)

knitr::kable(
  ogrenci_notlari_style,
  format = "html",
  escape = FALSE,
  caption = "Table 1: Student Scores Colored by Average"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```
In this table, red values show scores below the average, while green values show scores equal to or above the average.

```{r}
knitr::kable(
  head(temiz_penguinler, 10),
  caption = "Table 2: First 10 rows of temiz_penguinler"
)%>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

## Basic Statistical Analysis

```{r}
summary_table <- data.frame(
  Variable = c("Mathematics", "Physics", "Chemistry"),
  Count = c(
    length(ogrenci_notlari$matematik_notu),
    length(ogrenci_notlari$fizik_notu),
    length(ogrenci_notlari$kimya_notu)
  ),
  Mean = c(
    mean(ogrenci_notlari$matematik_notu),
    mean(ogrenci_notlari$fizik_notu),
    mean(ogrenci_notlari$kimya_notu)
  ),
  Median = c(
    median(ogrenci_notlari$matematik_notu),
    median(ogrenci_notlari$fizik_notu),
    median(ogrenci_notlari$kimya_notu)
  ),
  SD = c(
    sd(ogrenci_notlari$matematik_notu),
    sd(ogrenci_notlari$fizik_notu),
    sd(ogrenci_notlari$kimya_notu)
  ),
  Min = c(
    min(ogrenci_notlari$matematik_notu),
    min(ogrenci_notlari$fizik_notu),
    min(ogrenci_notlari$kimya_notu)
  ),
  Max = c(
    max(ogrenci_notlari$matematik_notu),
    max(ogrenci_notlari$fizik_notu),
    max(ogrenci_notlari$kimya_notu)
  )
)

knitr::kable(
  summary_table,
  digits = 2,
  caption = "Table 3: Summary Statistics of Student Scores"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```
This table summarizes the main statistical measures of student scores, including mean, median, standard deviation, minimum, and maximum values.


```{r}
penguin_summary <- data.frame(
  Variable = "Bill Length",
  Count = sum(!is.na(temiz_penguinler$bill_length_mm)),
  Mean = mean(temiz_penguinler$bill_length_mm, na.rm = TRUE),
  Median = median(temiz_penguinler$bill_length_mm, na.rm = TRUE),
  SD = sd(temiz_penguinler$bill_length_mm, na.rm = TRUE),
  Min = min(temiz_penguinler$bill_length_mm, na.rm = TRUE),
  Max = max(temiz_penguinler$bill_length_mm, na.rm = TRUE)
)

knitr::kable(
  penguin_summary,
  digits = 2,
  caption = "Table 4: Summary Statistics of Penguin Bill Length"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
    )
```

This table shows the summary statistics of penguin bill length, which helps to understand the distribution of the data.

### T-test Analysis

```{r}
t_result <- t.test(ogrenci_notlari$matematik_notu, mu = 70)
t_result
```
This test evaluates whether the average mathematics score is significantly different from 70.


## Visualization
```{r, fig.cap="Figure 1: Distribution of mathematics scores"}
hist(
  as.numeric(ogrenci_notlari$matematik_notu),
  breaks = 5,
  main = "Distribution of Mathematics Scores",
  xlab = "Score",
  col = "lightblue",
  border = "white"
)

```


```{r, fig.cap="Figure 2: Relationship between flipper length and body mass"}
ggplot(temiz_penguinler, aes(x = bill_length_mm, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Distribution of Bill Length by Species",
    x = "Bill Length (mm)",
    y = "Density"
  ) +
  theme_minimal()
```


The histogram shows the distribution of student scores, while the density plot illustrates how values are distributed across penguin species.


## Outlier Detection with Package Functions

In this section, the outlier detection functions in the package are used with student score data. The function is applied to `matematik_notu`, which is a numeric column from the `ogrenci_notlari` data frame.

```{r}
aykiri_deger_analizi(ogrenci_notlari$matematik_notu)
```

The `aykiri_deger_analizi()` function checks whether there are outlier values in the mathematics scores.

```{r}
tespit_et_outlier(ogrenci_notlari$matematik_notu)
```

The `tespit_et_outlier()` function also detects outlier values. These functions help users check the data before doing further statistical analysis.

If the user gives non-numeric input, the function may give a warning. Therefore, numeric data should be used for this analysis.

## Conclusion

In this vignette, we introduced the datasets included in the package and demonstrated basic statistical analysis such as mean, median, and t-test. We also showed how to visualize data and detect outliers using the functions provided in the package. This helps users understand how to apply the package in real data analysis.